Hierarchically modelling \emph{Kepler} dwarfs and subgiants to improve inference of stellar properties with asteroseismology

### 2020-10-29 paper/main.tex
Updated tables

\renewcommand{\arraystretch}{1.2}

@@

% Abstract of the paper <= 250 words
\begin{abstract}
    With recent advances in modelling stars using high-precision asteroseismology, the systematic effects associated with our assumptions of stellar helium abundance ($Y$) and the mixing-length theory parameter ($\mlt$) are becoming more important. We apply a new method to improve the inference of stellar parameters for a sample of \emph{Kepler} dwarfs and subgiants across a narrow mass range ($0.8 < M < \SI{1.2}{\solarmass}$). In this method, we include a statistical treatment of $Y$ and the $\mlt$. We develop a hierarchical Bayesian model to encode information about the distribution of $Y$ and $\mlt$ in the population, fitting a linear helium enrichment law including an intrinsic spread around this relation and normal distribution in $\mlt$. We test various levels of pooling parameters, with and without solar data as a calibrator. We are able to constrain $\mlt$ in the population with a mean of $1.90\substack{+0.10\\-0.09}$ with, and $1.74\substack{+0.08\\-0.07}$ without solar data. We find the gradient for the enrichment law, $\Delta Y / \Delta Z$ also varies with and without the solar data, obtaining $1.05\substack{+0.28\\-0.25}$ and $1.6 \substack{+0.45\\-0.42}$ respectively. While accounting for the uncertainty in $Y$ and $\mlt$, we are still able to report statistical uncertainties of 2.5 per cent in mass, 1.2 per cent in radius, and 12 per cent in age. Our method can also be applied to larger samples which will lead to improved constraints on both the population level inference and the star-by-star fundamental parameters. We show that more work is needed to study the validity of using the Sun as a calibrator for $Y$ and $\mlt$, perhaps by including asteroseismic signatures of helium abundance.
\end{abstract}

@@

% WJC - cut may exclude more metal-rich stars?
We removed more evolved stars from the APOKASC sample by cutting those with $\log g < \SI{3.8}{\dex}$. We then kept stars within 1-$\sigma$ of $-0.5 < \metallicity < \SI{0.5}{\dex}$ to remove metal-poor stars. Main sequence stars with $M \gtrsim \SI{1.2}{\solarmass}$ are understood to have a convective, hydrogen-burning core, with some dependence on the chemical composition and choice of stellar physics \citep{Appourchaux.Antia.ea2015}. Stellar models with a convective core require the treatment of extra stellar physics such as overshooting, which is beyond the scope of this work. Therefore, we keep only stars with masses within 1-$\sigma$ of \SIrange{0.8}{1.2}{\solarmass} from the preferred set of results of \citetalias{Serenelli.Johnson.ea2017}.

@@

We built a stellar model grid to use in training the ANN. The grid includes four independent model inputs: stellar mass ($M$), initial helium fraction ($Y_{\rm init}$), initial metallicity ($\mathrm{[M/H]}_\mathrm{init}$), and the mixing-length parameter ($\alpha_{\rm mlt}$).  Ranges and grid steps of the four model inputs are summarised in Table \ref{tab:grid}. We computed each stellar evolutionary track from the Hayashi line to the base of red-giant branch defined here by $\log g$ = 3.6 dex. We also computed evolutionary tracks with input values at the midpoint between points on the grid for validating the ANN.

@@

Theoretical stellar oscillations were calculated with the \textsc{GYRE} code (version 5.1), which was developed by \citet{Townsend.Teitler2013}. We computed radial modes (for $\ell$ = 0) by solving the adiabatic stellar pulsation equations with the structural models generated by \textsc{MESA}. We determined the asteroseismic large separation ($\dnu$) for each model with theoretical radial modes to avoid the systematic offset of the scaling relation. We derived $\Delta \nu$ with the approach given by \citet{White.Bedding.ea2011}, which is a weighted least-squares fit to the radial frequencies as a function of $n$.

We chose to ignore the well known, yet poorly characterised impact of modelled oscillation mode inaccuracies in the near-surface region of the star \citep{Kjeldsen.Bedding.ea2008, Ball.Gizon2014, Sonoi.Samadi.ea2015}. % This presents only a small effect when considering the average large frequency spacing, $\dnu$ and is beyond the scope of this paper.

@@

In this work, we considered a fully-connected ANN, where each neuron in layer $k-1$ is connected to every neuron in layer $k$. The output of the $i$-th neuron in layer $k$ is, 
%
\begin{equation}
    x_{i, k}=f_k(\boldsymbol{w}_{i, k} \cdot \boldsymbol{x}_{k-1} + b_{i, k})
\end{equation}
%
where $f_k$ is the \emph{activation} function for the $k$-th layer, $\boldsymbol{w}_{i, k}$ are the weights connecting all the neurons in layer $k-1$ to the current neuron, and $b_{i, k}$ is the bias. This generalises such that the output of the $k$-th layer is,
%
\begin{equation}
    \boldsymbol{x}_{k}=f_k(\boldsymbol{W}_{k} \cdot \boldsymbol{x}_{k-1} + \boldsymbol{b}_{k}),
\end{equation}
%
where $\boldsymbol{W}_k$ is the matrix of weights leading to all neurons in the $k$-th layer. For a regression ANN, we typically have a linear activation function applied to the output of the final layer. Therefore, the output of a network of $M$ hidden layers with initial input $X$ is,
%
\begin{equation}
    \boldsymbol{Y} = \boldsymbol{W}_{M} \cdot f_{M-1}(\dots f_1(\boldsymbol{W}_1 \cdot f_0(\boldsymbol{W}_{0} \cdot \boldsymbol{X} + \boldsymbol{b}_{0}) + \boldsymbol{b}_1) ) + \boldsymbol{b}_{M}
\end{equation}
%
We also restricted our configuration to an ANN with the same number of neurons, $N$ in each hidden layer. Hereafter, we refer to our choice of neurons per layer, $N$ and hidden layers, $M$ as the \emph{architecture}.

To fit the ANN, we used a set of training data, $\boldsymbol{D}_\mathrm{train} = \{(\boldsymbol{X}_1, \boldsymbol{Y}_1) \dots (\boldsymbol{X}_{N_\mathrm{train}}, \boldsymbol{Y}_{N_\mathrm{train}})\}$ comprising $N_\mathrm{train}$ input-output pairs. We split the training data into pseudo-random batches, $\boldsymbol{D}_\mathrm{batch}$ because this has been shown to improve ANN stability and computational efficiency \citep{Masters.Luschi2018}. The set of predictions made for each batch is evaluated with an error function, $E(\boldsymbol{D_\mathrm{batch}})$, also known as the \emph{loss} which quantifies the difference between the training data and predictions. We also considered an addition to the loss called \emph{regularisation} which helps reduce over-fitting \citep{Goodfellow.Bengio.ea2016}. During fitting, the weights are updated after each batch using an algorithm called the \emph{optimizer}, back-propagating the error with the goal of minimising the loss.

@@

During initial tuning, we found that having stellar age as an input was unstable, because it varied heavily with the other input parameters. We mitigated this by introducing an input to describe the fraction of time a star had spent in a given evolutionary phase, $f_\mathrm{evol}$. 
%
\begin{equation}
    f_\mathrm{evol} = \begin{cases}
        f_\mathrm{MS},\quad &f_\mathrm{MS} \leq 1\\
        1 + \frac{\tau\,-\,\tau_\mathrm{MS}}{\tau_{\log g=3.6}\,-\,\tau_\mathrm{MS}},\quad &f_\mathrm{MS} > 1
    \end{cases}
\end{equation}
%
where $\tau_{\log g=3.6}$ is the age of the star at the end of the track,
%
\begin{equation}
    f_\mathrm{MS} = \frac{\tau}{\tau_\mathrm{MS}},
\end{equation}
%
and $\tau_\mathrm{MS}$ is the MS lifetime. In other words, a star with $f_\mathrm{evol} \in (0, 1]$ is in its MS phase, burning hydrogen in its core, and $f_\mathrm{evol} \in (1, 2]$ has left the MS. Consequently, $f_\mathrm{evol}$ gives the ANN information about the internal state of the star which affects the output observables. Otherwise, $f_\mathrm{evol}$ is a meaningless parameter, although it could loosely be interpreted as a measure of the evolutionary phase of the star.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/network_10.png}
    \caption{An artificial neural network comprising $M$ hidden layers with $N$ neurons per layer. Arrows connecting the nodes represent tunable weights.}
    \label{fig:net}
\end{figure}

@@

We found that the optimal choice of $N$ and $M$ varied depending on our choice of other ANN parameters. Therefore, each time we explored a new parameter, we trained an ANN with a grid of $(N,M)$ ranging from $(32, 2)$ to $(512, 10)$.

@@

After extensive tuning, we opted for an ANN with $N=128$ neurons in each of $M=6$ hidden layers. Each of the hidden layers used an ELU activation function and L2 weight regularisation with $\lambda = \num{1e-6}$. We trained the ANN for \num{50000} epochs with a \num{500} training data batches each containing \num{15472} input-ouptut pairs. To fit the ANN, we used an SGD optimiser with an initial learning rate of \num{1e-4} and momentum of \num{0.999} with an MAE loss function. Training took $\sim \SI{48}{\hour}$ on an NVidia Tesla V100 graphics processing unit (GPU).

@@

When we added the Sun to the pooled models, PPS and MPS, we obtained lower values of $\Delta Y / \Delta Z$ with increased precision. In both models, the resulting $\Delta Y / \Delta Z$ of approximately \numrange{0.8}{1.0} was consistent with the initial helium fraction expected from solar models with our choice of \citet{Asplund.Grevesse.ea2009} abundances \citep{Serenelli.Basu2010}. However, such solar models have been shown to not recover helioseismic measurements of helium in the Sun \citep{Basu.Antia2004, Serenelli.Basu.ea2009, Villante.Serenelli.ea2014}. Solar models with the older \citet{Grevesse.Sauval1998} abundances typically yield higher helium fractions more in-line with helioseismology. Interestingly, the $\Delta Y / \Delta Z$ from the PP and MP models are higher than those with the Sun, in closer agreement to those calibrated with the older abundances. We would need to extend our model to predict surface helium abundances and compare these with asteroseismic measurements indicators of helium to test whether this difference is significant.

@@

The PP model (without the Sun) favoured a mean mixing-length parameter of $\mu_\alpha \simeq 1.7$. Whereas, the PPS model yielded a higher value of $\mu_\alpha \simeq 1.9$. We found this was attributed to the addition of the Sun. The individual solar results for the PPS model yielded a value of $\mlt_\odot = 2.11\pm0.03$ which was considerably higher than the $\mlt$ obtained for the other stars in the sample (see Appendix \ref{sec:sun-res}). The solar value also exceeds typical solar calibrated values of $\sim 1.9$ for the same stellar evolution code \citep{Stancliffe.Fossati.ea2016}. The reason for this is unclear, except that the difference we see between between $\alpha_{\mathrm{MLT},\odot}$ and the rest of the sample is not unique. Previous work on the LEGACY sample of \emph{Kepler} dwarfs found the best fitting $\mlt$ for their sample approximately 90 per cent of the solar calibrated value for one of their pipelines with similar model physics to this work \citep{SilvaAguirre.Lund.ea2017}.

>>>>

\renewcommand{\arraystretch}{1.2}  % So row spacing is big enough
% \setlength{\tabcolsep}{5pt}  % So that results tables fit

@@

% Abstract of the paper <= 250 words
\begin{abstract}
    With recent advances in modelling stars using high-precision asteroseismology, the systematic effects associated with our assumptions of stellar helium abundance ($Y$) and the mixing-length theory parameter ($\mlt$) are becoming more important. We apply a new method to improve the inference of stellar parameters for a sample of \emph{Kepler} dwarfs and subgiants across a narrow mass range ($0.8 < M < \SI{1.2}{\solarmass}$). In this method, we include a statistical treatment of $Y$ and the $\mlt$. We develop a hierarchical Bayesian model to encode information about the distribution of $Y$ and $\mlt$ in the population, fitting a linear helium enrichment law including an intrinsic spread around this relation and normal distribution in $\mlt$. We test various levels of pooling parameters, with and without solar data as a calibrator. We are able to constrain $\mlt$ in the population with a mean of $1.90\substack{+0.10\\-0.09}$ with, and $1.74\substack{+0.08\\-0.07}$ without solar data. We find the gradient for the enrichment law, $\Delta Y / \Delta Z$ also varies with and without the solar data, obtaining $1.05\substack{+0.28\\-0.25}$ and $1.60 \substack{+0.45\\-0.42}$ respectively. While accounting for the uncertainty in $Y$ and $\mlt$, we are still able to report statistical uncertainties of 2.5 per cent in mass, 1.2 per cent in radius, and 12 per cent in age. Our method can also be applied to larger samples which will lead to improved constraints on both the population level inference and the star-by-star fundamental parameters. We show that more work is needed to study the validity of using the Sun as a calibrator for $Y$ and $\mlt$, perhaps by including asteroseismic signatures of helium abundance.
\end{abstract}

@@

% WJC - cut may exclude more metal-rich stars?
We removed more evolved stars from the APOKASC sample by cutting those with $\log g < \SI{3.8}{\dex}$. We then kept stars within 1-$\sigma$ of $-0.5 < \metallicity < \SI{0.5}{\dex}$ to remove metal-poor and -rich stars. Main sequence stars with $M \gtrsim \SI{1.2}{\solarmass}$ are understood to have a convective, hydrogen-burning core, with some dependence on the chemical composition and choice of stellar physics \citep{Appourchaux.Antia.ea2015}. Stellar models with a convective core require the treatment of extra stellar physics such as overshooting, which is beyond the scope of this work. Therefore, we keep only stars with masses within 1-$\sigma$ of \SIrange{0.8}{1.2}{\solarmass} from the preferred set of results of \citetalias{Serenelli.Johnson.ea2017}.

@@

We built a stellar model grid to use in training the ANN. The grid includes four independent model inputs: stellar mass ($M$), initial helium fraction ($Y_{\rm init}$), initial metallicity ($\mathrm{[M/H]}_\mathrm{init}$), and the mixing-length parameter ($\alpha_{\rm mlt}$).  Ranges and grid steps of the four model inputs are summarised in Table \ref{tab:grid}. We increased the resolution at higher $\metallicity$ to give a more consistent resolution in $Z$ which is used as an ANN input in Section \ref{sec:ann}. We computed each stellar evolutionary track from the Hayashi line to the base of red-giant branch defined here by $\log g$ = 3.6 dex. We also computed evolutionary tracks with input values at the midpoint between points on the grid for validating the ANN.

@@

Theoretical stellar oscillations were calculated with the \textsc{GYRE} code (version 5.1), which was developed by \citet{Townsend.Teitler2013}. We computed radial modes (for $\ell$ = 0) for 42 radial orders by solving the adiabatic stellar pulsation equations with the structural models generated by \textsc{MESA}. We determined the asteroseismic large separation ($\dnu$) for each model with theoretical radial modes to avoid the systematic offset of the scaling relation. We derived $\Delta \nu$ with the approach given by \citet{White.Bedding.ea2011}, which is a weighted least-squares fit to the radial frequencies as a function of $n$.

We chose to ignore the well known, yet poorly characterised impact of modelled oscillation mode inaccuracies in the near-surface region of the star \citep{Kjeldsen.Bedding.ea2008, Ball.Gizon2014, Sonoi.Samadi.ea2015}. This typically presents only a small effect when considering the average large frequency spacing, $\dnu$. % This may need more

@@

In this work, we considered a fully-connected ANN, where each neuron in layer $k-1$ is connected to every neuron in layer $k$. The output of the $i$-th neuron in layer $k$ is, 
%
\begin{equation}
    x_{i, k}=f_k(\boldsymbol{w}_{i, k} \cdot \boldsymbol{x}_{k-1} + b_{i, k})
\end{equation}
%
where $f_k$ is the \emph{activation} function for the $k$-th layer, $\boldsymbol{w}_{i, k}$ are the weights connecting all the neurons in layer $k-1$ to the current neuron, and $b_{i, k}$ is the bias. This generalises such that the output of the $k$-th layer is,
%
\begin{equation}
    \boldsymbol{x}_{k}=f_k(\boldsymbol{W}_{k} \cdot \boldsymbol{x}_{k-1} + \boldsymbol{b}_{k}),
\end{equation}
%
where $\boldsymbol{W}_k$ is the matrix of weights leading to all neurons in the $k$-th layer. For a regression ANN, we typically have a linear activation function applied to the output of the final layer. Therefore, the output of a network of $H$ hidden layers with initial input $X$ is,
%
\begin{equation}
    \boldsymbol{Y} = \boldsymbol{W}_{H} \cdot f_{H-1}(\dots f_1(\boldsymbol{W}_1 \cdot f_0(\boldsymbol{W}_{0} \cdot \boldsymbol{X} + \boldsymbol{b}_{0}) + \boldsymbol{b}_1) ) + \boldsymbol{b}_{M}
\end{equation}
%
We also restricted our configuration to an ANN with the same number of neurons, $N$ in each hidden layer. Hereafter, we refer to our choice of neurons per layer, $N$ and hidden layers, $H$ as the \emph{architecture}.

To fit the ANN, we used a set of training data, $\boldsymbol{D}_\mathrm{train} = \{(\boldsymbol{X}_1, \boldsymbol{Y}_1) \dots (\boldsymbol{X}_{N_\mathrm{train}}, \boldsymbol{Y}_{N_\mathrm{train}})\}$ comprising $N_\mathrm{train}$ input-output pairs. We split the training data into pseudo-random batches, $\boldsymbol{D}_\mathrm{batch}$ because this has been shown to improve ANN stability and computational efficiency \citep{Masters.Luschi2018}. The set of predictions made for each batch is evaluated with an error function, $E(\boldsymbol{D}_\mathrm{batch})$, also known as the \emph{loss} which quantifies the difference between the training data and predictions. We also considered an addition to the loss called \emph{regularisation} which helps reduce over-fitting \citep{Goodfellow.Bengio.ea2016}. During fitting, the weights are updated after each batch using an algorithm called the \emph{optimizer}, back-propagating the error with the goal of minimising the loss.

@@

During initial tuning, we found that having stellar age as an input was unstable, because it varied heavily with the other input parameters. We mitigated this by introducing an input to describe the fraction of time a star had spent in a given evolutionary phase, $f_\mathrm{evol}$. 
%
\begin{equation}
    f_\mathrm{evol} = \begin{cases}
        f_\mathrm{MS},\quad &f_\mathrm{MS} \leq 1\\
        1 + \frac{\tau\,-\,\tau_\mathrm{MS}}{\tau_\mathrm{end}\,-\,\tau_\mathrm{MS}},\quad &f_\mathrm{MS} > 1
    \end{cases}
\end{equation}
%
where $\tau_\mathrm{end}$ is the age of the star at the end of the track,
%
\begin{equation}
    f_\mathrm{MS} = \frac{\tau}{\tau_\mathrm{MS}},
\end{equation}
%
and $\tau_\mathrm{MS}$ is the MS lifetime. In other words, a star with $f_\mathrm{evol} \in (0, 1]$ is in its MS phase, burning hydrogen in its core, and $f_\mathrm{evol} \in (1, 2]$ has left the MS. Consequently, $f_\mathrm{evol}$ gives the ANN information about the internal state of the star which affects the output observables. Otherwise, $f_\mathrm{evol}$ is a meaningless parameter, although it could loosely be interpreted as a measure of the evolutionary phase of the star.

@@

\begin{figure}
    \includegraphics[width=\linewidth]{figures/network_10.png}
    \caption{An artificial neural network comprising $H$ hidden layers with $N$ neurons per layer. Arrows connecting the nodes represent tunable weights.}
    \label{fig:net}
\end{figure}

@@

We found that the optimal choice of architecture ($N$ and $H$) varied depending on our choice of other ANN parameters. Therefore, each time we explored a new parameter, we trained an ANN with a grid of $(N,H)$ ranging from $(32, 2)$ to $(512, 10)$.

@@

After extensive tuning, we opted for an ANN with $N=128$ neurons in each of $H=6$ hidden layers. Each of the hidden layers used an ELU activation function and L2 weight regularisation with $\lambda = \num{1e-6}$. We trained the ANN for \num{50000} epochs with a \num{500} training data batches each containing \num{15472} input-output pairs. To fit the ANN, we used an SGD optimiser with an initial learning rate of \num{1e-4} and momentum of \num{0.999} with an MAE loss function. Training took $\sim \SI{48}{\hour}$ on an NVidia Tesla V100 graphics processing unit (GPU).

@@

When we added the Sun to the pooled models, PPS and MPS, obtained $\Delta Y / \Delta Z$ up to 2-$\sigma$ lower than the models without the Sun. In both models, the resulting $\Delta Y / \Delta Z$ of approximately \numrange{0.8}{1.0} was consistent with the initial helium fraction expected from solar models with our choice of \citet{Asplund.Grevesse.ea2009} abundances \citep{Serenelli.Basu2010}. However, such solar models have been shown to not recover helioseismic measurements of helium in the Sun \citep{Basu.Antia2004, Serenelli.Basu.ea2009, Villante.Serenelli.ea2014}. Solar models with the older \citet{Grevesse.Sauval1998} abundances typically yield higher helium fractions more in-line with helioseismology. Interestingly, the $\Delta Y / \Delta Z$ from the PP and MP models are higher than those including the Sun, in closer agreement to those calibrated with the older abundances. We could extend our model to include asteroseismic indicators of helium to improve the uncertainties on $Y$ and test whether this difference becomes more significant.

@@

The PP model (without the Sun) favoured a mean mixing-length parameter of $\mu_\alpha \simeq 1.7$. Whereas, the PPS model yielded a higher value of $\mu_\alpha \simeq 1.9$ by $\sim 2$-$\sigma$. We found this was attributed to the addition of the Sun. The individual solar results for the PPS model yielded a value of $\mlt_\odot = 2.12\pm0.03$ which was considerably higher than the $\mlt$ obtained for the other stars in the sample (see Appendix \ref{sec:sun-res}). The solar value also exceeds typical solar calibrated values of $\sim 1.9$ for the same stellar evolution code \citep{Stancliffe.Fossati.ea2016}. The reason for this is unclear, except that the difference we see between between $\alpha_{\mathrm{MLT},\odot}$ and the rest of the sample is not unique. Previous work on the LEGACY sample of \emph{Kepler} dwarfs found the best fitting $\mlt$ for their sample approximately 90 per cent of the solar calibrated value for one of their pipelines with similar model physics to this work \citep{SilvaAguirre.Lund.ea2017}.
